---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

```{r}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
destfile <- "activity.zip"
if (!file.exists(destfile)) {
  download.file(url = url, destfile = destfile, method = "curl")
}

datafile <- "activity.csv"
if(!file.exists(datafile)) {
        unzip(destfile)
}

dataWithNA <- read.table(datafile, header = TRUE, sep = ",")

## Convert date and remove NA values
dataWithNA$date <- as.Date(as.character(dataWithNA$date), "%Y-%m-%d")
data <- dataWithNA[!is.na(dataWithNA$steps),]

head(data)
```

## What is mean total number of steps taken per day?

```{r}
library(plyr)

## Total steps by date
stepsSummary <- ddply(data, .(date), summarize, total=sum(steps))

```
### Mean steps taken each day
```{r}
mean(stepsSummary$total)
```
### Median steps taken each day
```{r}
median(stepsSummary$total)

## Histogram of totals steps per day
hist(stepsSummary$total, main = "Histogram of total steps per day", xlab = "Steps", col = "blue")
```

## What is the average daily activity pattern?

```{r}

library(plyr)
library(ggplot2)

## Mean steps by interval for every day of data collected
stepsSummary <- ddply(data, .(interval), summarize, avgSteps=mean(steps))

## Plot average steps per interval
plot(avgSteps ~ interval, stepsSummary, type = "l", 
     ylab = "Steps", xlab = "Interval", main = "Average Daily Activity")
```

####The 5-minute interval that, on average, contains the maximum number of steps

```{r}
## find interval with maximum average steps
index <- which.max(stepsSummary$avgSteps)
stepsSummary$interval[index]
```

## Imputing missing values

### Code to describe and show a strategy for imputing missing data

Missing values can skew the results. Let's first see how big a problem with have by calculating the number of NA values.

```{r}
naRows <- sum(is.na(dataWithNA$steps))
totalRows <- nrow(dataWithNA)
print("NA Rows:")
naRows
print("Total Rows:")
totalRows
print("Percentages of missing data:")
naRows / totalRows * 100
```
### Create a new dataset by imputing NA values
#### The code replaces NA values with mean values for the given interval.

```{r}

## Summarize with mean values per interval
stepsSummary <- ddply(data, .(interval), summarize, avgSteps=mean(steps))

## Replace NA values with mean value for interval
cleanData <- dataWithNA
for (row in 1:nrow(cleanData)){
        if (!is.na(cleanData$steps[row]))
                next()
        
        interval <- cleanData$interval[row]
        avgId <- which(stepsSummary$interval == interval)
        cleanData$steps[row] <- stepsSummary$avgSteps[avgId]
}



```
## What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
stepsSummaryNoNA <- ddply(cleanData, .(date), summarize, total=sum(steps))

stepsSummaryWithNA <- ddply(dataWithNA, .(date), summarize, total=sum(steps))

## Mean total steps by date imputed
mean(stepsSummaryNoNA$total)

## Mean totals steps
mean(stepsSummaryWithNA$total, na.rm=TRUE)

## Median total steps by date imputed
median(stepsSummaryNoNA$total)

## Median totals steps
median(stepsSummaryWithNA$total, na.rm=TRUE)

## Sum total steps imputed
sum(stepsSummaryNoNA$total)

## Sum totals steps
sum(stepsSummaryWithNA$total, na.rm=TRUE)
```

### Imputing produces a slight change in the median number of steps per day and total steps observed is increased significantly. In the histograms the observed frequency increases.

## Histogram of total steps by day Imputed vs Non-Imputed
```{r}
par(mfrow=c(1,2))
hist(stepsSummaryNoNA$total, main = "(Imputed) Steps per Day", xlab = "Steps", col = "blue")
hist(stepsSummaryWithNA$total, main = "(Non-Imputed) Steps per Day", xlab = "Steps", col = "blue")
```

## Are there differences in activity patterns between weekdays and weekends?

### Yes!

```{r}
library(lattice)
## Add new column with day name
cleanData$day <- weekdays(cleanData$date)
weekend <- c("Saturday", "Sunday")

cleanData$dayType <- factor(cleanData$day %in% weekend, 
       levels=c(FALSE, TRUE), labels=c('weekday', 'weekend'))

## Summarize weekend and weekday data mean steps per interval
stepsSummary <- ddply(cleanData, .(interval, dayType), 
                             summarize, avgSteps=mean(steps))

## Plot average steps per interval weekday vs weekend
xyplot(avgSteps ~ interval | dayType, data = stepsSummary, type = "l", ylab = "Steps", xlab = "Interval", main = "Weekend Versus Weekday Steps", layout = c(1,2))
```